Convergent Adaptation while adapting

### 2014-07-02 patchy-selection-paper.tex

The convergent evolution of similar phenotypes in response to shared
selection pressures is a testiment to the power of selection to
repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes \citep{zhen2012parallel}. 
Such precise convergence points to the conservation of function of the genes underlying adaptive traits, 
sometimes over deep time scales \citep{deephomologypapers}. 
Convergence on a molecular level also suggests that the path 
accessible to adaptation is sometimes relatively constrained \citep{stern2013genetic}, 
as would be the case if only a few changes confer the advantagous phenotype 
and are sufficiently free of deleterious pleiotropic consequences.

@@

Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genotypic changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes REFs.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{dmitri}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
and standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{macnair1991evolution,schat1996identical,turner2010serpentine,mimulus};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches. 
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes \citep{benson1933concealing}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is primarily dark on a number of black lava flows separated by much lighter soil \citep{dice1940ecologic}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible from the dark pigmentation adaptation on different outcrops \citep{nachman2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism}
and in lizards \citep{whitesands}.

In this paper, we study this situation, namely,
when a set of alleles provide an adaptive benefit in geographically localized patches, 
separated by (inhabited) regions where the alleles are deleterious.
\plr{or neutral?}
The main questions are:
Under what conditions is it likely that each patch adapts in parallel, through new mutation,
and when is it likely that migration carries these alleles between patches?
How could we tell, after the fact, which has occurred?
We work in a model of continuous geography,
using a variety of known results and new methods.
In section XX we \dots QUICK OUTLINE

@@

\paragraph{Past work on spatial models}
We will make use of several previous results from the literature on
migration in spatially varying selection models. 
\citet{haldane1948theory}, \citet{fisher1950frequencies}, and later \citet{slatkin1973geneflow} first analyzed
deterministic models of of this sort.
\citet{nagylaki1975conditions} and \citet{conley1975application} showed that in one dimension, if the physical width of the patch is less than $(\sigma/\sqrt{2s_p}) \tan^{-1} (\sqrt{s_m/s_p})$, 
then there is no stable equilibrium with both $b$ and $B$ present,  % see p.603 in Nagylaki
so that migrational swamping prevents $B$ from establishing \citep[see also][ for a review]{lenormand2002limits}.
\citet{barton1987establishment}, using general theory of \citet{pollak1966survival}, showed  
that patches must be at least this critical size so that a new mutation has positive probability of establishment
(in the large population density limit).
The same paper also showed that
this probability of establishment of a new mutation decays exponentially with distance from the patch, 
with the scale given by $\sigma/\sqrt{s_m}$. Mutations appearing within the patch have probability of establishment
strictly less than the probability for a panimictic population,
but approaching this as the size of the patch increases.
This latter panmictic probability of establishment we denote $p_e$,
and often approximate by $2 s_p / \xi^2$ \citep{haldane1927mathematical,fisher1930genetical}.
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{maruyama1970fixation,cherry2003diffusion}.

@@

Finally we can consider the expected length of haplotype shared within a patch. 
To do this, we compare the rate of recombination between B and b haplotypes (equation~\ref{eqn:rate-rec-in-patch})
to the rate of coalescence of B haplotypes within a patch. 
We will assume that mixing within the patch is fast compared to the rate of drift, 
so that the rate of coalescence of B alleles within a patch is 
$\xi^2/(A\rho)$,
the ``variance effective population size'' of the patch
(this bypasses the well-known difficulties to spatial coalescence \citep{felsenstein1978pain,bartonetheridgedepaulis}).
The rate of recombination and coalescence will be on the same order if
$\xi^2 /(A\rho) \approx r (\sigma/\sqrt{s}) A^{-1/d}$, 
and hence the genetic length of haplotype shared between pairs of
individuals within the patches is on the order of
\begin{align}
  r = \frac{\sqrt{s} \xi^2}{\sigma \rho A^{(d-1)/d}} 
\end{align}
which does not depend on $A$ in one dimension,
but decreases with $\sqrt{A}$ in two dimensions.

>>>>

The convergent evolution of similar phenotypes in response to shared
selection pressures is a testiment to the power of selection to
repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes \citep{zhen2012parallel}. 
Such precise convergence points to the conservation of function of the genes underlying adaptive traits, 
sometimes over deep time scales \citep{zhen2012parallel,martin2013repeated,deephomologypapers}. 
Convergence on a molecular level also suggests that the path 
accessible to adaptation is sometimes relatively constrained \citep{stern2013genetic}, 
as would be the case if only a few changes confer the advantagous phenotype 
and are sufficiently free of deleterious pleiotropic consequences.

@@

Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genotypic changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes REFs.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
and standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{macnair1991evolution,schat1996identical,turner2010serpentine};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches \citep{wright2013indirect}.
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes \citep{benson1933concealing}.
\plr{we are missing what Nachman said was the original ref for this... I think you took a picture?}
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is primarily dark on a number of black lava flows separated by much lighter soil \citep{dice1940ecologic}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible from the dark pigmentation adaptation on different outcrops \citep{nachman2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism}
and in lizards \citep{rosenblum2010molecular}.

In this paper, we study this situation, namely,
when a set of alleles provide an adaptive benefit in geographically localized patches, 
separated by (inhabited) regions where the alleles are deleterious.
The main questions are:
Under what conditions is it likely that each patch adapts in parallel, through new mutation,
and when is it likely that migration carries these alleles between patches?
How could we tell, after the fact, which has occurred?
We work in a model of continuous geography,
using a variety of known results and new methods.
In section XX we \dots QUICK OUTLINE

@@

\paragraph{Past work on spatial models}
We will make use of several previous results from the literature on
migration in spatially varying selection models. 
\citet{haldane1948theory}, \citet{fisher1950frequencies}, and later \citet{slatkin1973geneflow} first analyzed
deterministic models of of this sort.
\citet{nagylaki1975conditions} and \citet{conley1975application} showed that in one dimension, if the physical width of the patch is less than $(\sigma/\sqrt{2s_p}) \tan^{-1} (\sqrt{s_m/s_p})$, 
then there is no stable equilibrium with both $b$ and $B$ present,  % see p.603 in Nagylaki
so that migrational swamping prevents $B$ from establishing \citep[see also][ for a review]{lenormand2002limits}.
\citet{barton1987establishment}, using general theory of \citet{pollak1966survival}, showed  
that patches must be at least this critical size so that a new mutation has positive probability of establishment
(in the large population density limit).
The same paper also showed that
this probability of establishment of a new mutation decays exponentially with distance from the patch, 
with the scale given by $\sigma/\sqrt{s_m}$. Mutations appearing within the patch have probability of establishment
strictly less than the probability for a panimictic population,
but approaching this as the size of the patch increases.
This latter panmictic probability of establishment we denote $p_e$,
and often approximate by $2 s_p / \xi^2$ \citep{haldane1927mathematical,fisher1930genetical}.
This result holds quite generally for a geographically spread population that experiences a uniform selection
pressure \citep{maruyama1970fixation,cherry2003diffusion}. 
More general work on the mathematically equivalent problem of density-dependent population growth 
has obtained much more general criteria
under which a habitat configuration will support a stable polymorphic equilibrium \citep{cantrell1991diffusive},
and determined the habitat configuration that maximizes the probability of establishment \citep{lou2006minimization}.

@@

Finally we can consider the expected length of haplotype shared within a patch. 
To do this, we compare the rate of recombination between B and b haplotypes (equation~\ref{eqn:rate-rec-in-patch})
to the rate of coalescence of B haplotypes within a patch. 
We will assume that mixing within the patch is fast compared to the rate of drift, 
so that the rate of coalescence of B alleles within a patch is 
$\xi^2/(A\rho)$,
the ``variance effective population size'' of the patch
(this bypasses the well-known difficulties to spatial coalescence \citep{felsenstein1975pain,barton2002neutral}).
The rate of recombination and coalescence will be on the same order if
$\xi^2 /(A\rho) \approx r (\sigma/\sqrt{s}) A^{-1/d}$, 
and hence the genetic length of haplotype shared between pairs of
individuals within the patches is on the order of
\begin{align}
  r = \frac{\sqrt{s} \xi^2}{\sigma \rho A^{(d-1)/d}} 
\end{align}
which does not depend on $A$ in one dimension,
but decreases with $\sqrt{A}$ in two dimensions.
